---
title: "AppliedBiostats"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## This is a code book for the second half of the 2025 Experimental design and applied biostatistics course PT7005

---
title: "Applied Biostats"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Simple linear regression practicum

Does age associate with blood pressure?

```{r}
#$load in the data
save(bpf, file = "bpf.Rdata")
load("bpf.Rdata")

# load libraries
library(ggplot2)
library(tidyverse)
library(skimr)
```

## Visualizing you data

```{r}
# a smart way to get a global look at your dataframe
skim(bpf)

# lets do some individual plots

# Create a box plot
p1 <- ggplot(bpf, aes(x = gender, y = bpf, fill = gender)) +
  geom_boxplot(alpha = 0.7) +  # Box plot with transparency
  geom_jitter(width = 0.2, alpha = 0.3) +  # Adds jittered points for visualization
  labs(title = "Box Plot of bpf by Gender",
       x = "Gender",
       y = "bpf") +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "salmon"))  # Custom colors
p1

# Create a scatter plot
p2 <- ggplot(bpf, aes(x = age, y = bpf)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Age vs bpf",
       x = "Age",
       y = "bpf") +
  theme_minimal()
p2

# Plot a histogram of a single continuous variable 
p3 <- ggplot(bpf, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Age",
       x = "Age",
       y = "Count") +
  theme_minimal()
p3

# Plot a histrogram now overlay gender
p4 <- ggplot(bpf, aes(x = age, fill = gender)) +
  geom_histogram(binwidth = 5, position = "dodge", color = "black") +
  labs(title = "Side-by-Side Histogram of Age by Gender",
       x = "Age",
       y = "Count",
       fill = "gender") +
    geom_density(linewidth = 1) +  # Add density curves
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "salmon"))
p4

ggplot(bpf, aes(x = age, fill = gender, color = gender)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, position = "identity", alpha = 0.4) +
  geom_density(size = 1) +  # Add density curves
  labs(title = "Histogram of Age by Gender with Density Curves",
       x = "Age",
       y = "Density",
       fill = "gender",
       color = "gender") +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "salmon")) +  # Custom fill colors
  scale_color_manual(values = c("blue", "red"))  # Custom line colors


plot_multi_histogram <- function(df, feature, label_column) {
    plt <- ggplot(df, aes(x=eval(parse(text=feature)), fill=eval(parse(text=label_column)))) +
    geom_histogram(alpha=0.7, position="identity", aes(y = ..density..), color="black") +
    geom_density(alpha=0.7) +
    geom_vline(aes(xintercept=mean(eval(parse(text=feature)))), color="black", linetype="dashed", size=1) +
    labs(x=feature, y = "Density")
    plt + guides(fill=guide_legend(title=label_column))
}
p4 <- plot_multi_histogram(bpf, 'bpf', 'gender')
p4


library(patchwork)
(p1 | p2) / (p3 | p4)
(p1 | p4)
```

## Running a simple linear regression

```{r}
# correlation
#cor.test(bpf$age, bpf$bpf)
#simple linear regression
lmodel <- lm(bpf ~ age, data = bpf)

#lmodel <- lm(Y ~ x1, data = mreg) # build linear model with Y outcome and x1 predictor and our dataframe mreg

library(performance) # load performance package to check assumptions
check_model(lmodel) # check assumptions of linear reg
summary(lmodel)

# go back to powerpoint

#check assumptions 
library(easystats)
check_model(lmodel) #output in console
model_performance(lmodel)
confint(lmodel) # calculate confidence intervals

# Make some nice plots
library(ggstatsplot)
ggscatterstats(data = mreg, x = x1, y = Y)
ggscatterstats(data = bpf, x = age, y = bpf)

```

## Confidence interval visual

```{r}
# Load libraries
library(ggplot2)

# Generate data
set.seed(123)
df <- data.frame(
  x = rnorm(100, mean = 50, sd = 10)
)
df$y_sig <- 2 * df$x + rnorm(100, sd = 10)  # Strong relationship (significant)
df$y_nonsig <- rnorm(100, mean = 50, sd = 30)  # No real relationship (non-significant)

# Plot significant and non-significant regression lines
ggplot(df, aes(x = x)) +
  geom_point(aes(y = y_sig), color = "blue", alpha = 0.6) +
  geom_smooth(aes(y = y_sig), method = "lm", color = "blue", fill = "lightblue") +
  geom_point(aes(y = y_nonsig), color = "red", alpha = 0.6) +
  geom_smooth(aes(y = y_nonsig), method = "lm", color = "red", fill = "pink") +
  labs(title = "Comparison of Significant vs. Non-Significant Confidence Intervals",
       x = "X Variable", y = "Y Variable") +
  theme_minimal()

sigmodel <- lm(y_sig ~ x, data = df)
nonsigmodel <- lm(y_nonsig ~ x, data = df)
summary(nonsigmodel)
confint(nonsigmodel)
confint(sigmodel)
```

## Standard Error

**The numerator — SE of the regression** This is the noise in your data. If your data points scatter wildly around the line, your residuals are big, your SE_regression is big, and consequently your SE(β̂₁) gets bigger too. More noise = less confidence in your slope.

**The denominator — spread of X values** This is √Σ(Xᵢ − X̄)², which measures how spread out your X values are. This is the more interesting and counterintuitive part:

-   If your X values are **tightly clustered** (everyone is nearly the same age, for example), the denominator is small → SE(β̂₁) is **large** — you can't see the slope well because you're only looking at a tiny slice of X

-   If your X values are **widely spread** (ages ranging from 20 to 80), the denominator is large → SE(β̂₁) is **small** — you have a much clearer view of how Y changes across X

## When is an outlier and outlier?

```{r}
# Load required libraries
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate X variable
n <- 100
x <- rnorm(n, mean = 50, sd = 10)

# Generate Y variable (Linear Relationship)
y <- 2 * x + rnorm(n, mean = 0, sd = 15)

# Introduce Two Influential Outliers
x_outliers <- c(20, 90)  # Extreme X values
y_outliers <- c(400, -200)  # Extreme Y values

# Append to original data
x_with_outliers <- c(x, x_outliers)
y_with_outliers <- c(y, y_outliers)

# Create data frames
df_no_outliers <- data.frame(x, y)
df_with_outliers <- data.frame(x = x_with_outliers, y = y_with_outliers)

# Fit Linear Models
lm_no_outliers <- lm(y ~ x, data = df_no_outliers)
lm_with_outliers <- lm(y ~ x, data = df_with_outliers)

summary(lm_with_outliers)
summary(lm_no_outliers)

check_model(lm_no_outliers) #output in console
check_model(lm_with_outliers) #output in console

check_outliers(
  lm_with_outliers,
  method = "cook",
  type = "default",
  iterations = 100,
  alternative = "two.sided"
)


# Compute Cook's Distance
cooks_d <- cooks.distance(lm_with_outliers)
cooks_d

# Identify points with Cook's D > 4/n (common threshold)
influential_points <- which(cooks_d > (4 / length(y_with_outliers)))

# Plot 1: Regression Without Influential Outliers
p1 <- ggplot(df_no_outliers, aes(x = x, y = y)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", fill = "lightblue") +
  labs(title = "Regression Without Influential Outliers", x = "X Variable", y = "Y Variable") +
  theme_minimal()

# Plot 1: Regression Without Influential Outliers
p2 <- ggplot(df_with_outliers, aes(x = x, y = y)) +
  geom_point(color = "red", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", fill = "lightblue") +
  labs(title = "Regression Without Influential Outliers", x = "X Variable", y = "Y Variable") +
  theme_minimal()

# Plot 2: Regression With Influential Outliers
p2 <- ggplot(df_with_outliers, aes(x = x, y = y)) +
  geom_point(color = "red", alpha = 0.6) +
  geom_point(data = df_with_outliers[influential_points, ], aes(x = x, y = y), color = "black", size = 4, shape = 8) +  # Mark influential points
  geom_smooth(method = "lm", color = "black", fill = "pink") +
  labs(title = "Regression With Influential Outliers (Cook's D Marked)", x = "X Variable", y = "Y Variable") +
  theme_minimal()

p2
# Print plots side by side
(p1 | p2)
```

## t-statistic

```{r}
# Load necessary library
library(ggplot2)

# Set degrees of freedom (df = n - 2, assume n = 20 for this example)
df <- 18  

# Generate t-distribution curve
t_values <- seq(-4, 4, length = 1000)
density_values <- dt(t_values, df)

# Define our t-statistic (example: t = 2.5)
t_stat <- 2.5

# Compute critical region (p-value area)
p_value_area <- 2 * pt(-abs(t_stat), df)  # Two-tailed test

# Create dataframe for plotting
data <- data.frame(t_values, density_values)

# Plot the t-distribution
ggplot(data, aes(x = t_values, y = density_values)) +
  geom_line(color = "blue", size = 1) +  # Plot t-distribution curve
  geom_area(data = subset(data, t_values > abs(t_stat)), 
            aes(x = t_values, y = density_values), fill = "red", alpha = 0.5) +  # Upper tail
  geom_area(data = subset(data, t_values < -abs(t_stat)), 
            aes(x = t_values, y = density_values), fill = "red", alpha = 0.5) +  # Lower tail
  geom_vline(xintercept = c(-abs(t_stat), abs(t_stat)), linetype = "dashed", color = "black") +  # Mark t-statistic
  labs(title = "t-Distribution with Critical Regions",
       subtitle = paste("t-statistic =", t_stat, "| p-value =", round(p_value_area, 4)),
       x = "t-value", y = "Density") +
  theme_minimal()

```

Degrees of freedom

```{r}
# Load necessary library
library(ggplot2)

# Generate t-distribution curves for different degrees of freedom
df_values <- c(5, 10, 20, 50)  # Different df values to compare
x <- seq(-4, 4, length = 1000)  # Range of t-values

# Create a dataframe for ggplot
data <- data.frame(
  x = rep(x, length(df_values)),
  y = unlist(lapply(df_values, function(df) dt(x, df))),
  df = factor(rep(df_values, each = length(x)))  # Convert df to a factor for color mapping
)

# Plot the t-distributions with different df
ggplot(data, aes(x, y, color = df)) +
  geom_line(size = 1) +
  stat_function(fun = dnorm, color = "black", linetype = "dashed", size = 1) + # Normal distribution
  labs(
    title = "Comparison of t-Distributions with Different Degrees of Freedom",
    subtitle = "Dashed black line represents the standard normal distribution",
    x = "t-value",
    y = "Density"
  ) +
  scale_color_manual(values = c("red", "blue", "green", "purple")) +
  theme_minimal()


```

## Heteroscedasticity

We are going to generate a dataframe with two continuous variables X and Y, where the Y dependent variable has an increasing variance with increasing levels of X. We will see this plotted on a residual vs fitted plot (where the residuals should show no discernible pattern and the reference line should be more or less flat and horizontal).

We will then plot this and visualize on a scatter plot and run the model and look at the SE, the t-stat and the p value. Which we will then compare to a model where. is the same but Y does not have increasing variance with increasing levels of X - what we should see if that because the t-stat is directly related to the SE, and the p value is a function of the t-distribution, we should see larger SEs a lower t-stat and a larger p-value (unreliable) in the model with heteroscedasticity.

```{r}
# Set seed for reproducibility
set.seed(123)

# Generate independent variable X
X <- rnorm(200, mean = 50, sd = 10)  # Continuous X
# Generate dependent variable Y with increasing variance
errors <- rnorm(200, mean = 0, sd = X / 5)  # Residuals increase with X
Y <- 5 + 2 * X + errors  # Linear relationship with heteroscedasticity
# Convert to dataframe
data <- data.frame(X, Y)

# Load ggplot2 for visualization
library(ggplot2)

# Plot Residuals vs. Fitted Values to visualize heteroscedasticity
model <- lm(Y ~ X)
ggplot(data, aes(model$fitted.values, resid(model))) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Plot (Heteroscedasticity)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Scatter plot with trend line
ggplot(data, aes(X, Y)) +
  geom_point(color = "blue", alpha = 0.6) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear fit
  labs(title = "Heteroscedasticity Example: Increasing Variance in Residuals",
       x = "X (Independent Variable)",
       y = "Y (Dependent Variable)") +
  theme_minimal()

summary(model)

library(lmtest)
bptest(model)
```

comparison with data that do not violate the assumption of homoscedasticity

```{r}
# Set seed for reproducibility
set.seed(123)

# Generate independent variable X
X <- rnorm(200, mean = 50, sd = 10)  # Continuous X

# Generate dependent variable Y with increasing variance
#errors <- rnorm(200, mean = 0, sd = X / 5)  # Residuals increase with X
#Y <- 5 + 2 * X + errors  # Linear relationship with heteroscedasticity

# Generate Y as a linear function of X with constant variance
errors <- rnorm(200, mean = 0, sd = 5)  # Constant variance error term
Y <- 5 + 2 * X + errors  # Linear relationship without heteroscedasticity

# Convert to dataframe
data <- data.frame(X, Y)

# Scatter plot with trend line
ggplot(data, aes(X, Y)) +
  geom_point(color = "blue", alpha = 0.6) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear fit
  labs(title = "Heteroscedasticity Example: Increasing Variance in Residuals",
       x = "X (Independent Variable)",
       y = "Y (Dependent Variable)") +
  theme_minimal()

# Plot Residuals vs. Fitted Values to visualize heteroscedasticity
model <- lm(Y ~ X)
ggplot(data, aes(model$fitted.values, resid(model))) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Plot (Heteroscedasticity)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

summary(model)
```

To overcome heteroscedasticity one could Use log transform

```{r}

# Generate dependent variable Y with increasing variance
errors <- rnorm(200, mean = 0, sd = X / 5)  # Residuals increase with X
Y <- 5 + 2 * X + errors  # Linear relationship with heteroscedasticity
Ylog = log(Y)

model <- lm(Ylog ~ X)
summary(model)

# Scatter plot with trend line
ggplot(data, aes(X, Ylog)) +
  geom_point(color = "blue", alpha = 0.6) +  # Scatter plot
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear fit
  labs(title = "Heteroscedasticity Example: Increasing Variance in Residuals",
       x = "X (Independent Variable)",
       y = "Y (Dependent Variable)") +
  theme_minimal()

# Plot Residuals vs. Fitted Values to visualize heteroscedasticity
model <- lm(Ylog ~ X)
ggplot(data, aes(model$fitted.values, resid(model))) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Plot (Heteroscedasticity)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

library(sandwich)
coeftest(model, vcov = vcovHC(model, type = "HC1"))

```

## Non-normality of residuals

```{r}
# Load necessary packages
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate independent variable X (continuous)
X <- rnorm(200, mean = 50, sd = 10)

# Generate Y with a skewed error term (non-normal residuals)
errors <- rexp(200, rate = 0.2) - 5  # Exponential distribution (right-skewed)
Y <- 5 + 2 * X + errors  # Linear relationship with non-normal residuals

# Fit a simple linear regression model
model <- lm(Y ~ X)

# Extract residuals
residuals <- resid(model)

# Q-Q Plot to check normality of residuals
ggplot(data.frame(residuals), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +  # Reference line for normality
  labs(title = "Q-Q Plot of Residuals (Non-Normal)",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

```

## Non-independence

Here we create two datasets, one with independent samples and another where we introduce autocorrelation, where sets of Y value depend on the previous value of Y (clustered data for example).

We see that in 1, the residuals are normally distributed and evenly-ish spaced around the OLS line and the other has clustered points. You can tell just by looking at a plot per se that you have violated the independence assumption, that should come from **knowledge about your study design**.

```{r}
# Load required library
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate independent data
n <- 100
x_indep <- rnorm(n, mean = 50, sd = 10)  # X variable
y_indep <- 2 * x_indep + rnorm(n, mean = 0, sd = 15)  # Y variable (independent)

# Generate non-independent data (autocorrelated Y)
y_auto <- numeric(n)
y_auto[1] <- 2 * x_indep[1] + rnorm(1, mean = 0, sd = 15)  # First value
for (i in 2:n) {
  y_auto[i] <- 0.7 * y_auto[i - 1] + (2 * x_indep[i]) + rnorm(1, mean = 0, sd = 10)  # Autocorrelation
}

# Create independent scatter plot with regression line
p1 <- ggplot(data.frame(x_indep, y_indep), aes(x = x_indep, y = y_indep)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", fill = "lightblue") +
  labs(title = "Independent Samples: Linear Model", x = "X Variable", y = "Y Variable") +
  theme_minimal()

# Create autocorrelated scatter plot with regression line
p2 <- ggplot(data.frame(x_indep, y_auto), aes(x = x_indep, y = y_auto)) +
  geom_point(color = "red", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", fill = "pink") +
  labs(title = "Non-Independent (Autocorrelated) Samples", x = "X Variable", y = "Y Variable") +
  theme_minimal()

# Print plots side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)

(p1 | p2)


```

## Centering for intercept and multicolinearity

When we do not "center" X, the intercept represents the value of Y when X is = 0. If a zero X doesn't exits in your data (e.g., age) then the intercept is meaningless. But, we can center our X (and covariates in the case of multiple linear regression). This makes the intercept = to the value of Y when X is at its mean.

```{r}
# Generate Data
set.seed(123)
X <- rnorm(100, mean = 50, sd = 10)  # Single predictor (e.g., Age)
Y <- 2 + 0.5 * X + rnorm(100, mean = 0, sd = 5)  # Outcome

# Fit Model Without Centering
model1 <- lm(Y ~ X)
summary(model1)  # Intercept represents Y when X = 0

# Center X
X_centered <- X - mean(X)

# Fit Model With Centering
model2 <- lm(Y ~ X_centered)
summary(model2)  # Intercept now represents Y at the mean of X

```

# Linear regression with a dichotomous predictor

What assumption are we making when using a linear model with a binary predictor? (equal variance between groups - homogeneity of variance!). Here we see that both lm and a t-test that assumes equal variance results in the same t-stat and p value and mean difference (although R t.test does not output the mean difference). Remember, b0 is equal to Y at the value of your reference value (0, which = females), mean Y of males therefore is b0 + b1, and so b1 = mean difference between males and females. Knowing how you coded your binary variable (which is the ref) is fundamental in interpreting your output.

```{r}
# Run you model with binary predictor
dlmodel <- lm(bpf ~ gender, data = bpf)
summary(dlmodel) # summary output
confint(dlmodel) # cinfidence intervals

# Compare to a t-test with and without assuming eqaul variance
t.test(bpf ~ gender, data = bpf) # Welch's t-test that doesnt assume equal variance
t.test(bpf ~ gender, data = bpf, var.equal = TRUE) # Pooled variance t-test closer to assmption of equal variance from lm 
```

# Multiple Regression

## Visualizing your data

Lets make some data, visualize it using skim and then make some plots. What do we see with X1 and X2? What do we need to look out for?

```{r}
# Lets generate some data first 
set.seed(123)  # Set seed for reproducibility

# Generate correlated predictors x1 and x2
n <- 200  # Sample size
x1 <- rnorm(n, mean = 50, sd = 10)  # Predictor 1
x2 <- 0.5 * x1 + rnorm(n, mean = 0, sd = 5)  # Predictor 2, mildly correlated with x1

# Independent predictor x3
x3 <- rnorm(n, mean = 30, sd = 8)

# Categorical predictor x4 with two levels
x4 <- sample(c("A", "B"), size = n, replace = TRUE)

# Create outcome Y as a function of predictors + random noise
Y <- 5 + 0.7 * x1 - 0.5 * x2 + 0.3 * x3 + ifelse(x4 == "A", 2, -2) + rnorm(n, mean = 0, sd = 3)

# Combine into a dataframe
mreg <- data.frame(Y, x1, x2, x3, x4)

# Overview of data 
skim(mreg)

# correlation matrix 
# (more options @ https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)
library(corrplot)
p1 <- corrplot(cor(mreg[, c("x1", "x2", "x3", "Y")]), method = "ellipse", addCoef.col = 'black', type = 'upper')

# Create a box plot of the categorical varibale and the Y outcome
p2 <- ggplot(mreg, aes(x = x4, y = Y, fill = x4)) +
  geom_boxplot(alpha = 0.7) +  # Box plot with transparency
  geom_jitter(width = 0.2, alpha = 0.3) +  # Adds jittered points for visualization
  labs(title = "Box Plot of X4 and Y",
       x = "X",
       y = "Y") +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "salmon"))  # Custom colors
p2

# Create a scatter plot of x2 and Y
p3 <- ggplot(mreg, aes(x = x2, y = Y)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "X2 vs Y",
       x = "x2",
       y = "Y") +
  theme_minimal()
p3

# Create a scatter plot of x1 and x2 
p4 <- ggplot(mreg, aes(x = x1, y = x2)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "X1 vs X2",
       x = "x1",
       y = "x2") +
  theme_minimal()
p4


(p2) / (p3 | p4)

# simulate some more data where x2 is a confounder and reduces the effect of x1 on y when added to the model because x2 accounts for some of the varaince in y.

# Set seed for reproducibility
set.seed(123)

# Simulate data
n <- 500  # Sample size
X2 <- rnorm(n, mean=50, sd=10)  # Confounder
X1 <- 0.5 * X2 + rnorm(n, mean=0, sd=5)  # Predictor correlated with X2
Y  <- 2.0 * X1 + 3.0 * X2 + rnorm(n, mean=0, sd=10)  # Outcome influenced by both X1 and X2

# Combine into a dataframe
mreg2 <- data.frame(Y, X1, X2)

# visualize using skim
skim(mreg2)

```

## Running multiple linear regression with confounds

Because X2 is correlated with X1 and Y, the effect of X1 is reduced when we include X2 in the model, because we arew now partitioning the variance between X1 and X2, isolating the effect of each on Y.

```{r}

# Fit the model without controlling for X2
model1 <- lm(Y ~ X1, data = mreg2)
summary(model1)

# Fit the model controlling for X2
model2 <- lm(Y ~ X1 + X2, data = mreg2)
summary(model2)
```

Visualizing your model can be helpful to tell your story, but simple scatter plots only tell the effect of one predictor (x1) on the outcome (Y). So we can use partial residual plots to illustrate the effect of your predictor (independent variable) of interest on Y after accounting for other covariates in the model.

To do this we must first extract the residuals from the model (the error variance or the variance not explained by the combined effect of all the variables in the model), and then add back in the estimated effect of just the predictor of interest (the estimate/coefficient for X1). This then shows the isolated effect of x1 on Y after accounting for the variance explained by the other predictors.

```{r}
# Compute partial residuals for 'x1'after accounting for X2
mreg2$partial_resid <- residuals(model2) + coef(model2)["X1"] * mreg2$X1

# Scatter plot with ggplot2
ggplot(mreg2, aes(x = X1, y = partial_resid)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Partial Residual Plot for X1",
       x = "X1", y = "Partial Residuals") +
  theme_minimal()

```

## Multicollinearity

If two proposed or utilized independent variables in your model are highly co-linear then their effects can become indistinguishable or uninterpretable because the model can arbitrarily assign more weight to one vs the other, it can inflate the standard errors and cause the coefficient estimates to be unstable.

Because pairwise correlation is not the same thing as shared predictive variance two correlated predictors can lead to low VIF (meaning they do not result in multicollinearity), so lets simulate some data that does.

```{r}
# Load necessary library
library(car)  # For VIF calculation

# Simulate data
set.seed(123)
n <- 100
X1 <- rnorm(n, mean = 10, sd = 5)  
X2 <- X1 * 0.9 + rnorm(n, mean = 0, sd = 1)  # Strong correlation with X1 (r ≈ 0.9)
Y  <- X1 * 3 + X2 * 2 + rnorm(n, mean = 0, sd = 5)  # X1 and X2 both predict Y

#make into dataframe
mreg3 <- data.frame(Y, X1, X2)

# Fit regression model
model3 <- lm(Y ~ X1 + X2, data = mreg3)
summary(model3)
check_model(model3)

# Calculate VIF (over 10 means multicollinearity)
vif(model3)

```

## Bootstrapping

Bootstrapping is a method of resampling. Your sample if often never going to match the true population parameter and so bootstrapping takes repeated samples (often with replacement) and runs the model or calculates to 95% CI or the effect within each bootstrap sample. You can then look at the distribution of the effects across each resample to see how much variance there would be if you ran your model in different samples.

```{r}
# Load necessary packages
library(tidymodels)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Simulate data
n <- 100
data <- tibble(
  X = rnorm(n, mean = 5, sd = 2),  # Predictor
  Y = 3 + 1.5 * X + rnorm(n, sd = 2)  # Outcome with noise
)

# Create 1000 bootstrap resamples
boot_resamples <- bootstraps(data, times = 1000)

# Define a linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

# Fit model on each bootstrap sample and extract beta_1 (slope)
boot_results <- boot_resamples %>%
  mutate(
    model = map(splits, ~ fit(lm_spec, Y ~ X, data = analysis(.x))),
    beta_1 = map_dbl(model, ~ tidy(.x) %>% filter(term == "X") %>% pull(estimate))
  )

# Extract bootstrapped beta estimates
beta_1_vals <- boot_results$beta_1

# Compute 95% confidence interval (percentile method)
ci_95 <- quantile(beta_1_vals, probs = c(0.025, 0.975))

# Calculate mean beta_1
mean_beta_1 <- mean(beta_1_vals)

# Plot the distribution of bootstrapped beta_1 estimates with 95% CI
ggplot(data.frame(beta_1_vals), aes(x = beta_1_vals)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean_beta_1, color = "red", linetype = "dashed", size = 1.2, label="Mean") +
  geom_vline(xintercept = ci_95[1], color = "blue", linetype = "dashed", size = 1, label="Lower 95% CI") +
  geom_vline(xintercept = ci_95[2], color = "blue", linetype = "dashed", size = 1, label="Upper 95% CI") +
  labs(title = "Bootstrapped Distribution of Beta_1 with 95% CI",
       x = "Bootstrapped Beta_1 Estimates",
       y = "Frequency") +
  theme_minimal()


model <- lm(Y ~ X, data = data)
summary(model)
mean_beta_1

```

# Mediation

The mediation (indirect, denoted as “ACME”) effects are calculated as the average difference between the direct and total effect. That is, if the effect of X on Y is reduced significantly when adding M into the model, then it is interpreted as though the mediator explains some of the effect of X on Y.

In the mediation package, two linear regression models are fit (the mediator model- the effect of X on M + covariates and the outcome model – effect of X on Y including M and covariates). The models constitute the input into the mediation model to calculate the Direct effects (‘ADE’ which is the effect of X on Y and could be interpreted as for every 1-unit increase in EELA there is an x (“estimate”) increase in Y), the average causal mediation effect, which is the indirect effect of X on Y through M, which is calculated as the total effect (the reduction in the effect of X on Y when accounting for M), minus the direct effect.

Direct effect is the adjusted effect of X on Y holding M constant

Total effect is the unadjusted effect of X on Y without M

Indirect effect is the ACME or the amount of the total effect that is explained by the mediator (c path)

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(mediation)
library(tidymodels)

# simulate some data using dlpyr
n <- 300
mediation <- tibble(
  ELE = rnorm(n, mean = 50, sd = 10),  # Predictor
  educ = 0.6 * ELE + rnorm(n, mean = 0, sd = 5), # Mediator
  cog = 0.3 * ELE + 0.7 * educ + rnorm(n, mean = 0, sd = 5)  # Outcome with noise
)

# save the dataframe
save(mediation, file = "mediation.Rdata") # save the data

load(file = "mediation.Rdata")

library(skimr)
skim(mediation)

library(mediation)
set.seed(465)
model.m <- lm(educ ~ ELE, mediation)
model.y <- lm(cog ~ ELE + educ, mediation)
model.t <- lm(cog ~ ELE, mediation)
results <- mediate(model.m, model.y, treat = "ELE", mediator = "educ",
                   boot = TRUE, sims = 1000)
summary(results)
# run the models
#set.seed(356)
#library(mediation)
#model.M <- lm(educ ~ ELE, mediation)
#model.Y <- lm(cog ~ ELE + educ, mediation)
#results <- mediate(model.M, model.Y, treat='ELE', mediator='educ',
#                  boot=TRUE, sims=1000)
#summary(results)
#summary(model.M)
#summary(model.Y)

summary(lm(cog ~ ELE, data = mediation))
summary(lm(cog ~ ELE + educ, data = mediation))
```

Mediation is still regression and so your assumptions are still important.

# Moderation

Continuous moderator interaction beta = how much the X-Y slope changes for every 1-unit increase in M- remember, the X-Y slope itself represents for every 1-unit change in X there is an x unit change in Y, so the X\*M beta is bounded by the units of M and the units of Y are preserved - just like in a lm without an interaction, we can standardize the beta's to interpret them as SD change.

To breakdown the interaction term with a simple slopes analysis, sometimes called marginal effects - this is where we calculate the beta slope for each level of M and test the null hypothesis under each level of M. If M is continuous, we can break it down into +1sd, -1sd and mean values of M, or if M has a unit that is easily interpretable, we can provide the range of M values for which to test the simple slopes (e.g., if Age is M, we can ask what the X-Y slope is for those age 10-20, 21-40, 41-60 ect).

For a dichotomous M, the beta estimate represents the difference in the X-Y slopes for each level of M.

```{r}
# Load required packages
library(ggplot2)
library(dplyr)

# Set seed for reproducibility
set.seed(123)

# Sample size
n <- 300

# Simulate categorical predictor (Intervention: 0 = Control, 1 = Treatment)
group <- sample(c(0, 1), size = n, replace = TRUE)

# Simulate continuous moderator (M1)
age <- rnorm(n, mean = 50, sd = 10)  # Continuous moderator (e.g., age, baseline score)

# Simulate categorical moderator (M2) with 3 levels
exercise <- factor(sample(c("Low", "Medium", "High"), size = n, replace = TRUE))

# Assign numerical values to categorical moderator for interaction effect
exercise_num <- as.numeric(exercise)  # Converts to 1 (Low), 2 (Medium), 3 (High)

# Generate outcome variable Y with main effects and interactions
Cognition <- 50 + 
  10 * group +       # Main effect of intervention
  0.5 * age +     # Main effect of continuous moderator
  5 * exercise_num +   # Main effect of categorical moderator
  3 * group * age +   # Interaction between X and M1
  4 * group * exercise_num + # Interaction between X and M2
  rnorm(n, mean = 0, sd = 5)  # Random noise

# Create a dataframe
moderation <- data.frame(group = factor(group, labels = c("Control", "Treatment")), age, exercise, Cognition)

save(moderation, file = "moderation.Rdata")

# View first few rows
head(moderation)

#convert categorical predictors to dummy vars
moderation <- moderation %>%
  mutate(exercise = as.numeric(exercise))

######
library(tidyverse)
set.seed(123)  # For reproducibility

# Number of observations
n <- 200  

# Simulate continuous predictor (X)
X <- rnorm(n, mean = 50, sd = 10)

# Simulate continuous moderator (m1)
m1 <- rnorm(n, mean = 30, sd = 5)

# Simulate categorical moderator (m2) with 3 levels
m2 <- sample(c("low", "medium", "high"), size = n, replace = TRUE)

# Convert categorical variable to factor
m2 <- factor(m2, levels = c("low", "medium", "high"))

# Generate outcome (Y) with interaction effects
# Base effect of X
Y <- 5 + 0.6 * X + 0.3 * m1 + 
     0.5 * (X * m1) +  # Interaction between X and continuous moderator
     ifelse(m2 == "medium", 0.8 * X, 0) +  # X effect changes at medium level
     ifelse(m2 == "high", -0.5 * X, 0) +   # X effect changes at high level
     rnorm(n, mean = 0, sd = 5)  # Random noise

# Create a tidy dataframe
df <- tibble(X, m1, m2, Y)

moderation <- df %>%
  rename(exercise = X,
         bmi = m1,
         educ = m2,
         cog = Y)
# View first few rows
print(head(moderation))

save(moderation, file = "moderation.Rdata")

load(file = "moderation.Rdata")

skim(moderation)


model2 <- lm(cog ~ exercise + bmi + exercise*bmi, moderation)
summary(model2)

interact_plot(model2, pred = exercise, modx = bmi, partial.residuals = TRUE)

simple_slopes(model2, pred = exercise, modx = bmi, control.fdr = TRUE, confint = TRUE)

#sim_slopes(model2, pred = exercise, modx = bmi, robust = "HC3", control.fdr = TRUE,   confint = TRUE)

# Build model including interacton term
#model2 <- lm(cog ~ exercise + bmi + exercise*bmi, data = moderation)
#summary(model2)
# Run interact plot
#library(interactions)
#interact_plot(model2, pred = exercise, modx = bmi, partial.residuals = TRUE)

# Calculate simple slopes
#library(ggstance)
#sim_slopes(model2, pred = exercise, modx = bmi, robust = "HC3", control.fdr = TRUE,   confint = TRUE)


# Calculate the level of age at the point which the slope between exercise and cognition becomes insignificant 
#johnson_neyman(model, pred = VO2max_rel, modx = BMI, mod2 = HouseholdIncome, alpha = .05)

```

# ANOVA and LM for categorical questions

In this chunk we are going to compare the use of ANOVA and linear regression to ask if there are differences in means of a continuous outcome between groups

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Simulate data: 3 groups with different means
set.seed(123)
df <- data.frame(
  group = rep(c("NZ", "UK", "USA"), each = 10),
  y = c(rnorm(10, mean = 5, sd = 1),  # Group A
        rnorm(10, mean = 6, sd = 1),  # Group B
        rnorm(10, mean = 7, sd = 1))  # Group C
)

# Convert 'group' to a factor
df$group <- as.factor(df$group)

anova <- df

save(anova, file = "anova.Rdata")

skim(anova)

# Visualize the data
ggplot(anova, aes(x = group, y = y, fill = group)) +
  geom_boxplot() +
  theme_minimal() +
  geom_point() +
  labs(title = "Boxplot of Groups", y = "Response Variable (y)")

# ANOVA Analysis
anova_model <- aov(y ~ group, data = anova)
summary(anova_model)
tukey_posthoc <- TukeyHSD(anova_model)
tukey_posthoc

# Linear Regression with Factor Coding
lm_model <- lm(y ~ group, data = anova)
summary(lm_model)

# re-order to make USA reference level
anova$group <- relevel(anova$group, ref = "USA")

# If your hypothesis wants you to ask for the difference in GPD between two specific pairs, then you can do pairwsie compariusoons, again contrlling for the number of comparisons you run. 

library(emmeans)
emmeans(lm_model, pairwise ~ group, adjust = "tukey")


```

# Repeated measures - two time points

We are going to simulate some data of two groups (intervention and control), two time points (pre and post) and a continuous outcome Y. We are going to then test the hypothesis that the intervention has a greater effect on the outcome compared to the control group, using 1. ANCOVA, 2. Repeated Measures ANOVA 3. Linear model.

```{r}
set.seed(123)  # For reproducibility

# Define sample size per group
n_per_group <- 50  # 50 participants per group

# Create data frame with ID, Group, and Time
data <- data.frame(
  ID = rep(1:(2 * n_per_group), each = 2),  # Each participant appears twice (Pre & Post)
  Group = rep(c("Control", "Treatment"), each = n_per_group * 2),  # Assign groups
  Time = rep(c("Pre", "Post"), times = 2 * n_per_group)  # Two time points per subject
)

# Generate baseline (Pre) scores
pre_scores <- rnorm(2 * n_per_group, mean = 50, sd = 10)  # Random normal distribution

# Generate Post scores with different changes for Control and Treatment groups
post_scores <- pre_scores + ifelse(data$Group == "Control", 
                                   rnorm(2 * n_per_group, mean = 2, sd = 5),  # Small change
                                   rnorm(2 * n_per_group, mean = 7, sd = 5))  # Larger change

# Assign scores based on time
data$Score <- ifelse(data$Time == "Pre", pre_scores, post_scores)

# View first few rows
head(data)





# ClAUDE DATA SIM
 
# R code to simulate data for a repeated measures study with treatment and control groups
# The dataset will include demographic variables and show greater pre-post change in the treatment group

# Set seed for reproducibility
set.seed(42)

# Set parameters
n_participants <- 100  # Total number of participants
prop_treatment <- 0.5  # Proportion in treatment group

# Demographic parameters
age_mean <- 35
age_sd <- 10
prop_female <- 0.55
education_mean <- 14
education_sd <- 3

# Outcome parameters
pre_score_mean <- 25
pre_score_sd <- 5
# Effect sizes
control_effect <- -2  # Small improvement in control group
treatment_effect <- -8  # Larger improvement in treatment group
post_score_sd <- 6

# Generate participant IDs
participant_id <- 1:n_participants

# Assign participants to groups (0=control, 1=treatment)
group <- sample(c(0, 1), n_participants, replace = TRUE, 
                prob = c(1-prop_treatment, prop_treatment))

# Generate demographic variables
age <- round(rnorm(n_participants, mean = age_mean, sd = age_sd))
# Ensure ages are reasonable
age <- pmax(18, pmin(75, age))

sex <- rbinom(n_participants, 1, prop_female)  # 1=female, 0=male

education <- round(rnorm(n_participants, mean = education_mean, sd = education_sd))
# Ensure education is reasonable
education <- pmax(8, pmin(22, education))

# Generate pre-treatment scores
pre_score <- rnorm(n_participants, mean = pre_score_mean, sd = pre_score_sd)

# Generate post-treatment scores
# Control group: small improvement
post_score_control <- pre_score[group == 0] + 
                      rnorm(sum(group == 0), mean = control_effect, sd = post_score_sd)

# Treatment group: larger improvement
post_score_treatment <- pre_score[group == 1] + 
                        rnorm(sum(group == 1), mean = treatment_effect, sd = post_score_sd)

# Combine post scores
post_score <- numeric(n_participants)
post_score[group == 0] <- post_score_control
post_score[group == 1] <- post_score_treatment

# Create the dataset
treatment_study <- data.frame(
  participant_id = participant_id,
  group = group,
  age = age,
  sex = sex,
  education = education,
  pre_score = pre_score,
  post_score = post_score
)

# Calculate change scores to verify effect
treatment_study$change <- treatment_study$post_score - treatment_study$pre_score

# Verify the data
summary_by_group <- aggregate(
  cbind(age, education, sex, pre_score, post_score, change) ~ group, 
  data = treatment_study, 
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
print(summary_by_group)

# Run t-test to verify the difference in change scores
t_test_result <- t.test(change ~ group, data = treatment_study)
print(t_test_result)

# Check if the effect is as expected
cat("\nMean change in control group:", mean(treatment_study$change[treatment_study$group == 0]))
cat("\nMean change in treatment group:", mean(treatment_study$change[treatment_study$group == 1]))

# Write the dataset to a CSV file
write.csv(treatment_study, "treatment_study.csv", row.names = FALSE)

# Let's also create some visualizations to check the data

# If you want to use ggplot2 for better visualizations
# install.packages("ggplot2")
library(ggplot2)

# Plot pre and post scores by group
ggplot(treatment_study, aes(x = factor(group), y = change, fill = factor(group))) +
  geom_boxplot() +
  labs(title = "Change Scores by Group",
       x = "Group (0 = Control, 1 = Treatment)",
       y = "Change Score (Post - Pre)",
       fill = "Group") +
  scale_fill_manual(values = c("lightblue", "salmon"),
                   labels = c("Control", "Treatment")) +
  theme_minimal()

# Create a summary dataset for pre/post visualization
library(reshape2)
treatment_study_long <- melt(treatment_study, 
                            id.vars = c("participant_id", "group"), 
                            measure.vars = c("pre_score", "post_score"),
                            variable.name = "time_point",
                            value.name = "score")

# Calculate means and standard errors by group and time point
summary_stats <- aggregate(
  score ~ group + time_point, 
  data = treatment_study_long, 
  FUN = function(x) c(mean = mean(x), se = sd(x)/sqrt(length(x)))
)

# Plot pre and post means with error bars
ggplot(summary_stats, aes(x = time_point, y = score[,"mean"], 
                          group = factor(group), color = factor(group))) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = score[,"mean"] - score[,"se"], 
                    ymax = score[,"mean"] + score[,"se"]),
                width = 0.2) +
  labs(title = "Pre and Post Scores by Group",
       x = "Time Point",
       y = "Score",
       color = "Group") +
  scale_color_manual(values = c("blue", "red"),
                    labels = c("Control", "Treatment")) +
  theme_minimal()
```

ANCOVA

```{r}
# Load necessary library
library(car)  # For ANCOVA diagnostics

# Reshape from long to wide format
ancova_data <- data %>%
  pivot_wider(names_from = Time, values_from = Score) %>%
  rename(Pre = Pre, Post = Post)  # Renaming for clarity

# View the first few rows
head(ancova_data)

# Fit ANCOVA model

# ANCOVA using AOV function
ancova_model_aov <- aov(Post ~ Group + Pre, data = ancova_data)

# summary of results
summary(ancova_model_aov)

# Run ANCOVA using lm
ancova_model <- lm(Post ~ Group + Pre , data = ancova_data)
summary(ancova_model)

# View summary of results
summary(ancova_model)

# Check ANCOVA assumptions (homogeneity of regression slopes)
Anova(ancova_model, type = "III")  # Type III SS for interaction check

```

Linear model

```{r}
# linear model 

lmod <- lm(Score ~ Group + Time + Group*Time, data = data)
summary(lmod)
```

# Repeated measures - 3+ time points

```{r}
# Load necessary libraries
library(lme4)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Create simulation parameters
n_subjects <- 50  # Number of subjects
n_timepoints <- 4  # Number of timepoints
time_values <- c(0, 1, 2, 3)  # Time points (e.g., months)
groups <- c("Control", "Treatment")  # Group conditions

# Set fixed effects
intercept <- 10  # Overall mean at baseline
group_effect <- 3  # Effect of being in treatment group
time_effect <- 1.5  # Effect of time (per unit)
interaction_effect <- 0.8  # Additional effect of time for treatment group

# Set random effects variance components
random_intercept_sd <- 2.0  # Between-subject variability in intercepts
random_slope_sd <- 0.7  # Between-subject variability in slopes
random_cor <- 0.3  # Correlation between random intercepts and slopes
residual_sd <- 1.5  # Within-subject variability (residual error)

# Create data frame
simdata <- data.frame()

# Generate data for each subject
for (i in 1:n_subjects) {
  # Assign group (alternating)
  group <- groups[(i %% 2) + 1]
  
  # Generate random effects for this subject
  # Using multivariate normal distribution for correlated random effects
  sigma <- matrix(c(random_intercept_sd^2, 
                    random_cor * random_intercept_sd * random_slope_sd,
                    random_cor * random_intercept_sd * random_slope_sd, 
                    random_slope_sd^2), 
                  nrow = 2)
  random_effects <- MASS::mvrnorm(1, mu = c(0, 0), Sigma = sigma)
  subject_intercept <- random_effects[1]
  subject_slope <- random_effects[2]
  
  # Generate observations for each timepoint
  for (t in 1:n_timepoints) {
    time <- time_values[t]
    
    # Calculate expected value based on fixed and random effects
    group_val <- ifelse(group == "Treatment", 1, 0)
    expected <- intercept + 
                group_effect * group_val + 
                time_effect * time + 
                interaction_effect * group_val * time +
                subject_intercept + 
                subject_slope * time
    
    # Add random noise (residual error)
    observed <- expected + rnorm(1, mean = 0, sd = residual_sd)
    
    # Add to data frame
    simdata <- rbind(simdata, data.frame(
      ID = paste0("S", i),
      Group = group,
      Time = time,
      Score = observed
    ))
  }
}

# Convert categorical variables to factors
simdata$ID <- factor(simdata$ID)
simdata$Group <- factor(simdata$Group)

# View the structure of the data
str(simdata)

# View a sample of the data
head(simdata, 10)


# subset
subset_data <- simdata %>%
  # Get distinct subject IDs, take first 60
  group_by(ID) %>%
  filter(ID %in% (unique(simdata$ID)[1:20]))

write.csv(subset_data, file = "mixedeffectsdata.csv")

```

```{r}
# fixed effects only model
lmodel <- lm(Score ~ Group * Time, data = subset_data)
# View model summary
summary(lmodel)

# random interceot only model
lmmodeli <- lmer(Score ~ Group * Time + (1|ID), data = subset_data)
# View model summary
summary(lmmodeli)

# Fit the mixed model with random intercept and slope
lmmodel <- lmer(Score ~ Group * Time + (1 + Time | ID), data = subset_data)
# View model summary
summary(lmmodel)

# Fit the mixed model with random intercept and slope assuming uncorrelated intercepts and slopes (WARNING)
model_uncorrelated <- lmer(Score ~ Group * Time + (1 + Time || ID), data = simdata)
summary(model_uncorrelated)

# Plot the data to visualize
ggplot(subset_data, aes(x = Time, y = Score, group = ID, color = Group)) +
  geom_line(alpha = 0.4) +
  geom_point(alpha = 0.4) +
  stat_summary(aes(group = Group), fun = mean, geom = "line", size = 1.5) +
  theme_minimal() +
  labs(title = "Simulated Longitudinal Data",
       subtitle = "Individual trajectories with group means",
       y = "Outcome Score", 
       x = "Time Point")
```